Matroids \& testing in families of algebraically dependent and contextually related hypotheses

### 2013-08-20 paper/draft.tex
added some reference, completed proof of weakly exhaustive set theorem

\title{Multiple testing with logical dependencies}

@@

\subsection{Pairwise comparisons of population means}
\begin{itemize}
\item All pairwise
\item Gold standard designs
\item Change point analysis
\item General treatment control - treatment treatment situations
\end{itemize}

\subsection{General linear contrast tests}
\begin{itemize}
\item Cost effectiveness (Georg)
\item SNP Studies (Hothorn-Hothorn)
\item Order restriction contrasts (Williams, ...)
\end{itemize}

@@

\begin{proof}
  Not yet done!
\end{proof}

To get a result on strictly exhaustive set we need to restrict
condition (\ref{cond:matroida}) further.

\begin{theorem}
  Consider a family of null hypotheses $\mathcal{H}$ for which above
  the more stringent conditions hold:
  \begin{align}
    \label{cond:matroid2}
    \bigcap_J \Theta_i &\nsubseteq \bigcup_{I\setminus J} \Theta_i,
    \text{ and} \\
    \bigcap_{J}\Theta_i &\nsupseteq     \bigcup_{I \setminus J} \Theta_i ,
  \end{align}
  % and additionally $\bigcap_S \Theta_i^C \neq
  % \emptyset$ (practically that means that all alternatives may be true
  % at the same time).
  Then independence as defined in Definition \ref{def:indep} gives a
  matroid for   which every flat corresponds to a strictly exhaustive
  index set and vice versa.
\end{theorem}

\begin{proof}
  Not yet done!
  % We first show by contradiction that every flat is strictly
  % exhaustive: Consider $I$ is a flat but that $I$ is not strictly
  % exhaustive. As $I$ is a flat, we have for every $l \in S\setminus
  % I$, $l \sim I$. Let $J \in \mathscr{I}$ be a maximally independent
  % set in $I$. Then $J \cup l \in \mathscr{I}$ and $|J \cup l| =
  % |J|+1$. If $I$ were not strictly exhaustive, then $\bigcap_{I}
  % \Theta_i \cap \bigcap_{S\setminus I} \Theta^c_i = \emptyset$. 
\end{proof}

\begin{comment}
  Above restriction (\ref{cond:matroid2}), already for the simple case
  of $|J|=1$, is no longer a statment about the parallels of
  $M$. 
\end{comment}

@@

The question remains how to compute $\mathcal{F}$. For this observe
that the $h_i$ define a hyperplane configuration and $\mathcal{F}$ is
just the face lattice of the corresponding oriented
matroid. Consequently, $\mathcal{F}$ can be found using algorithms
for arrangement construction from polyhedral programming
\cite{Avis-KF-92,Avis-KF-96,Ferrez-KF-Liebling-01} which are
polynomial in time and input size and compact in space!

>>>>

\title{Matroids \& testing in families of algebraically dependent and
  contextually related hypotheses}

@@

History 
\begin{itemize}
\item \cite{shaffer1986modified} represents pairwise comparisons
 by set partitions
\item \cite{bergmann1988improvements} define the concepts of an
 exhaustive sets
\item \cite{bernhard1991computergestuetzte} shows properties of
 exhaustive sets that are basically {\em closure axioms} of a matroid
 and provides algorithms to check each intersection hypotheses if is
 a flat using rank function.
\item \cite{westfall1997multiple} generates flats using dependence
 arguments in vector spaces.
\item \cite{weichert2000robuste} suggest an MTP based on clique
 decomposition of the graph, cliques in the graph are flats
\item \cite{westfall2007multiple} provide a branch and bound algorithm
 to generate all flats. They note that for pairwise comparisons all
 exhaustive hypotheses are {\em partition hypotheses}
\end{itemize}

\subsection{Pairwise comparisons of population means}
\begin{itemize}
\item All pairwise \citep{shaffer1986modified,weichert2000robuste}
\item Gold standard designs \citep{bauer1998testing,pigeot2003assessing}
\item Change point analysis (?)
\item General treatment control - treatment treatment situations 
\end{itemize}

\subsection{General linear contrast tests}
\begin{itemize}
\item Cost effectiveness \citep{haller2011cost,gutjahr2012multiple}
\item SNP Studies (Hothorn-Hothorn) \citep{hothorn2009order}
\item Order restriction contrasts (Williams, ...)
\end{itemize}

@@

\begin{proof}
  First assume that $I$ exhaustive, then for every $j \in S \setminus
  I$ $H_{I \cup j} \subsetneq H_{I}$, that is $j \nsim I$ for all $j
  \in S\setminus I$ and therefore $I$ is flat.

  Next, consider $I \subseteq S$ and $I \in \mathscr{L}$, that is
  $\sigma I = I$ assume that $I$ is not exhaustive, then $\exists J
  \supersetneq I$ with $H_I \subset H_J$. Therefore, $\exists j \in J
  \setminus I$ such that $H_I \in H_{J \setminus j}$ and therefore
  also $H_I \subseteq H_j$. Consequently $H_{I \cup j} = H_{I}$, that
  is $j \sim I$, which contradicts that $I$ is a flat and completes
  the proof.
\end{proof}

To get a result on strictly exhaustive set we need to restrict
condition (\ref{cond:matroida}) further. {\sl It is an open question
  whether one can find conditions that make it easier to show matroid
  properties}.

% \begin{theorem}
%   Consider a family of null hypotheses $\mathcal{H}$ for which above
%   the more stringent conditions hold:
%   \begin{align}
%     \label{cond:matroid2}
%     \bigcap_J \Theta_i &\nsubseteq \bigcup_{I\setminus J} \Theta_i,
%     \text{ and} \\
%     \bigcap_{J}\Theta_i &\nsupseteq     \bigcup_{I \setminus J} \Theta_i ,
%   \end{align}
%   % and additionally $\bigcap_S \Theta_i^C \neq
%   % \emptyset$ (practically that means that all alternatives may be true
%   % at the same time).
%   Then independence as defined in Definition \ref{def:indep} gives a
%   matroid for   which every flat corresponds to a strictly exhaustive
%   index set and vice versa.
% \end{theorem}

% \begin{proof}
%   Not yet done!
%   % We first show by contradiction that every flat is strictly
%   % exhaustive: Consider $I$ is a flat but that $I$ is not strictly
%   % exhaustive. As $I$ is a flat, we have for every $l \in S\setminus
%   % I$, $l \sim I$. Let $J \in \mathscr{I}$ be a maximally independent
%   % set in $I$. Then $J \cup l \in \mathscr{I}$ and $|J \cup l| =
%   % |J|+1$. If $I$ were not strictly exhaustive, then $\bigcap_{I}
%   % \Theta_i \cap \bigcap_{S\setminus I} \Theta^c_i = \emptyset$. 
% \end{proof}

% \begin{comment}
%   Above restriction (\ref{cond:matroid2}), already for the simple case
%   of $|J|=1$, is no longer a statment about the parallels of
%   $M$. 
% \end{comment}

@@

The question remains how to compute $\mathcal{F}$. For this observe
that the $h_i$ define a hyperplane configuration and $\mathcal{F}$ is
just the face lattice of the corresponding oriented
matroid. Consequently, $\mathcal{F}$ can be found using algorithms
for arrangement construction from polyhedral programming
\citep{avis1992pivoting,avis1996reverse,ferrez2001cuts} which are
polynomial in time and input size and compact in space!

\bibliographystyle{alpha}
\bibliography{cleanrefs}
